Parsimonious Representation of Nonlinear Dynamical Systems using Diffusion Maps: A Chemotaxis Case Study

### 2014-12-11 chemotaxis/paper/draft1.tex
moved old + new text to chemotaxis/detecting harmonics draft

\documentclass[prl, reprint]{revtex4-1}

\usepackage{amsmath}
\usepackage{amsfonts}

\usepackage{graphicx}
\usepackage{subcaption}

\usepackage{epstopdf}

@@

\title{Characterizing System Dynamics using Diffusion Maps: A Chemotaxis Case Study}

\author{Carmeline J. Dsilva}
\email{cdsilva@princeton.edu}
\affiliation{Department of Chemical and Biological Engineering, Princeton University, Princeton, NJ, 08544}

\author{Ronen Talmon}
\email{ronen.talmon@yale.edu}
\affiliation{Department of Mathematics, Yale University, New Haven, CT, 06520}

\author{Ronald R. Coifman}
\email{coifman@math.yale.edu}
\affiliation{Department of Mathematics, Yale University, New Haven, CT, 06520}

\author{Ioannis G. Kevrekidis}
\email{yannis@princeton.edu}
\affiliation{Department of Chemical and Biological Engineering, Princeton University, Princeton, NJ, 08544}
\affiliation{Program in Applied and Computational Mathematics, Princeton University, Princeton, NJ, 08544}

\date{\today}

@@

Many dynamical systems are governed by stochastic motions at the microscale which give rise to smooth, coherent dynamics at the macroscale.
%
The relationship between the microscopic and macroscopic dynamics is often not immediately obvious and either requires some modeling assumptions or extensive observation of the system.
%
We demonstrate automatically extracting a parameterization of microscale data using manifold learning techniques, and show that this parameterization is consistent with the governing macroscopic dynamical model. 
%
These data-driven algorithms need no {\em a priori} knowledge about the appropriate macroscopic model, but instead, only require the appropriate observers and pairwise metric between observations.
%
We also show how these algorithms can detect changes in the macroscopic dynamical behavior as a function of the microscopic system parameters. 
%
We illustrate these ideas using a model which arises in studies of cellular chemotaxis.

\end{abstract}

\keywords{diffusion maps, earth mover's distance, histograms}

\maketitle

%\section{Introduction}

@@

\begin{figure*}[t!]
\def\figwidth{6.4cm}
\begin{subfigure}{\figwidth}
\includegraphics[width=\textwidth]{EMD_withhist_p_1}
\caption{}
\label{subfig:small_lambda_p}
\end{subfigure}
\begin{subfigure}{\figwidth}
\includegraphics[width=\textwidth]{EMD_withhist_t_1}
\caption{}
\label{subfig:small_lambda_t}
\end{subfigure}
\begin{subfigure}{\figwidth}
\includegraphics[width=\textwidth]{EMD_withhist_p_400}
\caption{}
\label{subfig:large_lambda_p}
\end{subfigure}
\begin{subfigure}{\figwidth}
\includegraphics[width=\textwidth]{EMD_withhist_t_400}
\caption{}
\label{subfig:large_lambda_t}
\end{subfigure}
\caption{Diffusion maps embeddings computed from simulation data of the velocity jump process with (a,b) $\lambda=1$, $s=1$, and (c,d) $\lambda=400$, $s=20$.  For each set, we run many simulations, each with $N=1000$ particles.
%
We vary the initial conditions such that $0.1 \le p  \le 0.9$, and allow each simulation to evolve for $10$ time units.
%
We discard the initial portion of each simulation trajectory ($t < 1$), which corresponds to the initial relaxation of the system.
%
The observation $z(i) \in \mathbb{R}^n$ is the histogram of $x_1(i), x_2(i), \dots, x_N(i)$ into $n = 32$ bins.
%
The distances used in the diffusion maps kernel are the earth mover's distances between the histograms of particle positions. The data are colored by (a, c) $p$, the initial probability of a particle moving to the right, and (b, d) $t$, time. Representative histograms are shown for selected data points.}
\label{fig:dmaps_embed_emd}
\end{figure*}

@@

\begin{acknowledgments}

The authors would like to thank...

\end{acknowledgments}

\bibliography{../../references/references}

\end{document}
%\end{document}

>>>>

\documentclass[preprint]{elsarticle}

\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsfonts}
%\usepackage{cite}
%\usepackage{graphics}
%\usepackage{epsf}
%\usepackage{psfig}
%\usepackage{float}
\usepackage{graphicx}
\usepackage{epstopdf}
%\usepackage[small]{caption2}
%\usepackage{bm}%
%\usepackage{multirow}
%\usepackage{subfigure}
%\usepackage{wrapfig}
%\usepackage{color}
%\usepackage{dsfont}
%\usepackage{amsthm}
%\usepackage{algorithm}
%\usepackage{subfigure}
\usepackage{subcaption}

\graphicspath{{../}}

%\newtheorem{theorem}{Theorem}
%\newtheorem{corollary}[theorem]{Corollary}
%\newtheorem{lemma}[theorem]{Lemma}
%\newtheorem{observation}[theorem]{Observation}
%\newtheorem{proposition}[theorem]{Proposition}
%\newtheorem{definition}[theorem]{Definition}
%\newtheorem{claim}[theorem]{Claim}
%\newtheorem{fact}[theorem]{Fact}
%\newtheorem{assumption}[theorem]{Assumption}
%%\newtheorem{example}[theorem]{Example}
% \newtheorem{remark}[theorem]{Remark}

\journal{Computers \& Chemical Engineering}

@@

\begin{frontmatter}

\title{Parsimonious Representation of Nonlinear Dynamical Systems using Diffusion Maps: A Chemotaxis Case Study}

\author[princeton]{Carmeline~J.~Dsilva\corref{cor1}}
\ead{cdsilva@princeton.edu}
%
\author[technion]{Ronen~Talmon}
\ead{ ronen@ee.technion.ac.il}
%
\author[yale]{Ronald R. Coifman}
\ead{coifman@math.yale.edu}
%
\author[princeton, princetonpacm]{Ioannis~G.~Kevrekidis \corref{cor1}}
\ead{yannis@princeton.edu}
%
\address[princeton]{Department of Chemical and Biological Engineering, Princeton University, Princeton, NJ, 08540, USA}
\address[technion]{Technion - Israel Institute of Technology, Haifa, 3200003, Israel}
\address[yale]{Department of Mathematics, Yale University, New Haven, CT, 06520, USA}
\address[princetonpacm]{Program in Applied and Computational Mathematics, Princeton University, Princeton, NJ, 08540, USA}
%
\cortext[cor1]{Corresponding author}

@@

TODO
\end{abstract}

%\sloppy

\begin{keyword}
TODO
\end{keyword}

\end{frontmatter}

\section{Introduction}

\begin{itemize}

\item
Data driven analysis methods, and in particular, manifold learning, can do many things (dimensionality reduction, reduction in dynamical systems, etc).

\item
Manifold learning constructs parameterization of the data through the spectral analysis of a Laplace operator.

\item
Initially it was used almost exclusively for toy examples/synthetic data (``these people suck").

\item
Only recently, has it been applied to real (experimental/simulation) data, and this was possible due to advances in observers (data representation) and metrics (``we are awesome").

\item
However, a major limitation (of spectral methods based on a Laplace operator) is identifying the true principal components.

\item
In this paper, we propose an algorithm to automatically identify those principal components using locally linear regression.

\item
All of these considerations (recently developed plus what we propose here) provide a data analysis pipeline. We show that this pipeline can successfully analyze data from a complex stochastic dynamical system originated from the study of cellular dynamics (chemotaxis).

\end{itemize}

\section{Issues in Manifold Learning/Diffusion maps}

\subsection{Diffusion maps}

\subsection{Observers}
Add physical examples, and Amit's bi-spectrum, scattering....

Cite PNAS, stating that the histograms are good since they linearize any noise.

\subsection{Metric}
Euclidean (not good, especially for histograms), Mahalanobis (good, cite PNAS, since it noise robust). Here we use EMD since it is multiscale.
Stability and regularity issues.

\section{Identify the principal components}

\subsection{Algorithm: Local Linear Regression}

Given the eigenvectors $\phi_1, \phi_2, \dots, \phi_{n-1} \in \mathbb{R}^n$, we would like to automatically deduce which ones are harmonics/functions of another. 
%
To accomplish this, we will fit a function $f(\phi_1, \dots, \phi_k)$ to $\phi_{k+1}$. 
%
If the resulting fit is good/accurate, then $\phi_{k+1}$ is most likely a harmonic of $\phi_1, \dots, \phi_k$.

We will use a local linear function 
\begin{equation}
f_{k+1}(\phi_1, \dots, \phi_k) = a_0 + a_1 \phi_1 + \dots + a_k \phi_k
\end{equation}
where the coefficients $a_0, \dots, a_k$ are functions of $\phi_1, \dots, \phi_k$.

To find these coefficients, we want to minimize the local weighted sum of squares
\begin{equation} \label{eq:opt_problem}
\sum_{i=1}^n w_i(x) (\phi_{k+1}(x) - f_{k+1} (\phi_1(x), \dots, \phi_{k}(x))
\end{equation}
where $w$ is a weighting function.
%
We will use
%
\begin{itemize}
\item $w_i(x) = \exp \left( - \frac{\|\Phi_{k}(x) - \Phi_{k} (x_i) \|^2}{\epsilon^2} \right)$, where $\Phi_k(x) = \begin{bmatrix}
\phi_1(x) \\
\vdots \\
\phi_k(x)
\end{bmatrix} $
\item $\epsilon = M / 3$, where $M$ is the median of the pairwise distances between $\Phi_{k}(x)$
\end{itemize}

Let $\hat{f}_{k+1}(\phi_1(x_i), \dots, \phi_k(x_i))$ denote the minimum achieved in \eqref{eq:opt_problem}, when the minimization is performed over $x_j$ with $j \ne i$. 
%
We then define the leave-one-out- cross-validation error as
\begin{equation}
r_{k+1} = \sum_{i=1}^n \left( \phi_{k+1} (x_i) - \hat{f}_{k+1}(\phi_1(x_i), \dots, \phi_k(x_i)) \right)^2
\end{equation}
%
A small value of $r_{k+1}$ means that $\phi_{k+1}$ can be accurately approximated from $\phi_1, \dots, \phi_k$, and is therefore a harmonic of a previous mode.
%
A large value of $r_{k+1}$ indicates that $\phi_{k+1}$ parameterizes a new direction in the data.

\subsection{Eigenvalues to Quantify Dimensions}

The eigenvalues of the Laplacian with Neumann boundary conditions on a $d$-dimensional rectangular domain are given by
\begin{equation}
\mu_{k,l} = \frac{k^2 \pi^2}{L_l^2}
\end{equation}
for $k=1, 2, \dots$ and $l=1, 2, \dots, d$, and $L_l$ is the length of the domain along the $l^{th}$ dimension. 
%
These eigenvalues are related to the eigenvalues we compute in diffusion maps by 
\begin{equation}
\lambda_{k,l} = e^{-\mu_{k,l} t} = e^{-\frac{k^2 \pi^2}{L_l^2} t}
\end{equation}
%
Therefore, 
\begin{equation}
\sqrt{\frac{\log(\lambda_{1,l_1})}{\log(\lambda_{1,l_2})}} = 
\frac{L_{l_2}}{L_{l_1}}
\end{equation}
and so looking at the ratio of the log of the eigenvalues (modulo harmonics) should inform us as to the lengths of the different directions in the manifold.

\subsection{Illustrative Examples}

\subsubsection{Swiss Roll}

\includegraphics[height=3in]{swissroll1_evals}
\includegraphics[height=3in]{swissroll2}\\
\includegraphics[height=2in]{swissroll1_cv}
\includegraphics[height=2in]{swissroll2_cv}\\
\includegraphics[height=3in]{swissroll1_color1}
\includegraphics[height=3in]{swissroll2_color1}\\
\includegraphics[height=3in]{swissroll1_color2}
\includegraphics[height=3in]{swissroll2_color2}

For first case: $\lambda_1 = 0.9982$, $\lambda_2 = 0.9959$

For second case: $\lambda_1 = 0.9977$, $\lambda_5 = 0.9652$.

\subsubsection{Torus}

\section{Chemotaxis: a case study}

@@

\begin{figure}[t!]
\begin{subfigure}{6cm}
\includegraphics[width=\textwidth]{EMD_withhist_p_1}
\caption{}
\label{subfig:small_lambda_p}
\end{subfigure}
\begin{subfigure}{6cm}
\includegraphics[width=\textwidth]{EMD_withhist_t_1}
\caption{}
\label{subfig:small_lambda_t}
\end{subfigure}
\begin{subfigure}{6cm}
\includegraphics[width=\textwidth]{EMD_withhist_p_400}
\caption{}
\label{subfig:large_lambda_p}
\end{subfigure}
\begin{subfigure}{6cm}
\includegraphics[width=\textwidth]{EMD_withhist_t_400}
\caption{}
\label{subfig:large_lambda_t}
\end{subfigure}
\caption{Diffusion maps embeddings computed from simulation data of the velocity jump process with (a,b) $\lambda=1$, $s=1$, and (c,d) $\lambda=400$, $s=20$.  For each set, we run many simulations, each with $N=1000$ particles.
%
We vary the initial conditions such that $0.1 \le p  \le 0.9$, and allow each simulation to evolve for $10$ time units.
%
We discard the initial portion of each simulation trajectory ($t < 1$), which corresponds to the initial relaxation of the system.
%
The observation $z(i) \in \mathbb{R}^n$ is the histogram of $x_1(i), x_2(i), \dots, x_N(i)$ into $n = 32$ bins.
%
The distances used in the diffusion maps kernel are the earth mover's distances between the histograms of particle positions. The data are colored by (a, c) $p$, the initial probability of a particle moving to the right, and (b, d) $t$, time. Representative histograms are shown for selected data points.}
\label{fig:dmaps_embed_emd}
\end{figure}

@@

\subsection{Detect dimensionality}

\includegraphics[height=2in]{lle_errors1}
\includegraphics[height=2in]{lle_errors2}

\includegraphics[height=2in]{lle_evals1}
\includegraphics[height=2in]{lle_evals2}

\includegraphics[height=2in]{lle_embed1_1}
\includegraphics[height=2in]{lle_embed2_2}

\hfill 
\includegraphics[height=2in]{lle_embed1_2}

\section*{Acknowledgment}
The authors would like to thank TODO.

\bibliographystyle{elsarticle-num}
\bibliography{papers}

\end{document}
